---
title: "Data analysis of CSF proteomics"
author: "Matthijs de Geus"
output:
  html_document:
    toc: yes
    toc_float: yes
    code_folding: show
  '': default
---

# Load dependencies and datasets
```{r}
library(tidyverse)
library(caret)
library(ranger)
library(tidymodels)
library(gplots)
library(UpSetR)
library(cowplot)
library(ComplexHeatmap)
library(rpart.plot)
library(pROC)

protein_data_DEM_Apos <- read.csv("protein_data.csv")
```


# Linear regression modeling
```{r}
to_linearmodels <- protein_data_DEM_Apos 

to_linearmodels$Pathology_group <- factor(to_linearmodels$Pathology_group, levels = c("CU","AD","NPH"))
lm_1 <- lm(as.matrix(to_linearmodels[,6:ncol(to_linearmodels)]) ~ Pathology_group + Age + Gender_binary, data = to_linearmodels)
to_linearmodels$Pathology_group <- factor(to_linearmodels$Pathology_group, levels = c("AD","NPH"))
lm_2 <- lm(as.matrix(to_linearmodels[,6:ncol(to_linearmodels)]) ~ Pathology_group + Age + Gender_binary, data = to_linearmodels) 

tidy_lm1 <- broom::tidy(lm_1) %>%
  filter(term != "(Intercept)") %>%
  mutate(versus = "CU")
tidy_lm2 <- broom::tidy(lm_2) %>%
  filter(term != "(Intercept)") %>%
  mutate(versus = "AD")

tidy_res <- rbind(tidy_lm1, tidy_lm2) %>%
  rename("Protein" = response) %>%
  mutate(term = gsub("Pathology_group", "", term)) %>%
  arrange(p.value)

Dx_res <- tidy_res %>%
  filter(term == "CU" | term == "AD" | term == "NPH") %>%
  mutate(padj = p.adjust(p.value, method = "BH")) 

Dx_res_sig <- Dx_res %>%
  filter(padj <= 0.05) %>%
  unite("Contrast", c(term, versus), remove = FALSE) 

Dx_res_GSEA <- Dx_res %>%
  unite("Contrast", c(term, versus), remove = FALSE) %>%
  separate(Protein, into = c("Protein Name", "Uniprot_ID"), remove = FALSE, sep ="_") %>%
  separate(Uniprot_ID, into = c("Uniprot_ID"), sep = "\\.")
lm_GSEA_AD_CU <- Dx_res_GSEA %>% filter(Contrast == "AD_CU")
lm_GSEA_NPH_AD <- Dx_res_GSEA %>% filter(Contrast == "NPH_AD")
lm_GSEA_NPH_CU <- Dx_res_GSEA %>% filter(Contrast == "NPH_CU")
```

### Upset plot for linear model output
```{r}
lm_upset_data <- Dx_res_sig %>%
  unite("Contrast", term, versus, sep = " vs ", remove = FALSE)
AD_NPH_lm_sig <- lm_upset_data %>% filter(Contrast == "NPH vs AD")
CU_NPH_lm_sig <- lm_upset_data %>% filter(Contrast == "NPH vs CU")
CU_AD_lm_sig <- lm_upset_data %>% filter(Contrast == "AD vs CU")


lm_upset <- data.frame(Protein = c(unique(lm_upset_data$Protein))) %>%
  mutate(`CU vs AD` = ifelse(Protein %in% CU_AD_lm_sig$Protein, 1, 0)) %>%
  mutate(`CU vs NPH` = ifelse(Protein %in% CU_NPH_lm_sig$Protein, 1, 0)) %>%
  mutate(`AD vs NPH` = ifelse(Protein %in% AD_NPH_lm_sig$Protein, 1, 0)) %>%
  mutate(Sums = rowSums(.[2:4]))

upset(lm_upset, 
      order.by = "freq",
      matrix.color = "grey20",
      main.bar.color = "grey10",
      shade.color = "grey70",
      sets.x.label = "",
      mainbar.y.label = "Number of Proteins",
      text.scale = 1.4
      )

lm_three_contrasts <- lm_upset %>%
  filter(Sums == 3) %>%
  select(Protein) %>%
  separate(Protein, into = c("Protein_name", NA)) %>%
  pull(Protein_name)
```

### Volcano plots
```{r}
to_vp <- Dx_res %>%
  unite("contrast", term, versus , sep = "_", remove = FALSE) %>%
  # select(Protein, estimate, padj) %>%
  mutate(log_padj = -log10(padj))

vp_down <- to_vp %>%
  filter(padj < 0.05) %>%
  filter(estimate < 0)
vp_up <- to_vp %>%
  filter(padj < 0.05) %>%
  filter(estimate > 0)

Dx_res_sig_NPH <- Dx_res_sig %>%
  filter(Contrast == "NPH_AD" | Contrast == "NPH_CU") %>%
  separate(Protein, into = c("Protein_name", "UniprotID"), sep = "_")

ggplot(to_vp, aes(x = estimate, y = log_padj))+
  geom_point(aes(color = (padj < 0.05)))+
  scale_color_manual(values = c("grey", "blue")) +
  facet_wrap(~contrast)+
  geom_hline(aes(yintercept = 1.30103), linetype = "dashed") +
  labs(x = "Estimate", y = "-log10(p-value)", title = "Differentially expressed proteins") 
```

# Full sample decision tree
```{r}
# Create a decision tree model specification
tree_spec <- decision_tree() %>%
 set_engine("rpart") %>%
 set_mode("classification")

## Tree of all three classes
all_classes_to_tree <- protein_data_DEM_Apos %>%
  select(Pathology_group, Gender_binary, Age, c(6:ncol(protein_data_DEM_Apos))) %>%
  mutate(Pathology_group = droplevels(as.factor(Pathology_group)))

tree_fit_all <- rpart(Pathology_group ~ ., data = all_classes_to_tree, control = rpart.control(maxdepth = 7))
rpart.plot(tree_fit_all, type = 4, extra = 101, under = TRUE, cex = 0.6, box.palette = list("CU" = "#fee090", "AD" = "#4575b4", "NPH" = "#d73027"), roundint = FALSE)
```

# Random Forest models
```{r}
CU_AD <- protein_data_DEM_Apos %>%
  filter(Pathology_group == "CU" | Pathology_group == "AD") %>%
  select(Pathology_group, Gender_binary, Age, c(6:ncol(protein_data_DEM_Apos))) %>%
  mutate(Pathology_group = droplevels(as.factor(Pathology_group)))

CU_NPH <- protein_data_DEM_Apos %>%
  filter(Pathology_group == "CU" | Pathology_group == "NPH") %>%
  select(Pathology_group, Gender_binary, Age, c(6:ncol(protein_data_DEM_Apos))) %>%
  mutate(Pathology_group = droplevels(as.factor(Pathology_group)))

AD_NPH <- protein_data_DEM_Apos %>%
  filter(Pathology_group == "AD" | Pathology_group == "NPH") %>%
  select(Pathology_group, Gender_binary, Age, c(6:ncol(protein_data_DEM_Apos))) %>%
  mutate(Pathology_group = droplevels(as.factor(Pathology_group)))
```

### Define the tuning grids for each comparison
```{r}
myGrid_CU_AD <- data.frame(.mtry = seq(from = 1, to = ncol(CU_AD)-1, by = 20),
                     .splitrule = "gini",
                     .min.node.size = 1)

myGrid_CU_NPH <- data.frame(.mtry = seq(from = 1, to = ncol(CU_NPH)-1, by = 20),
                     .splitrule = "gini",
                     .min.node.size = 1)

myGrid_AD_NPH <- data.frame(.mtry = seq(from = 1, to = ncol(AD_NPH)-1, by = 20),
                     .splitrule = "gini",
                     .min.node.size = 1)
```

### 10-fold cross validation repeated 5 times
```{r}
set.seed(42)
# CU vs AD
model_CU_AD <- train(
  Pathology_group ~ .,
  CU_AD,
  method = "ranger",
  tuneGrid = myGrid_CU_AD,
  importance = "impurity",
  trControl = trainControl(
    method = "repeatedcv",
    number = 10,
    repeats = 5,
    verboseIter = TRUE
  )
)

# CU vs NPH
model_CU_NPH <- train(
  Pathology_group ~ .,
  CU_NPH,
  method = "ranger",
  tuneGrid = myGrid_CU_NPH,
  importance = "impurity",
  trControl = trainControl(
    method = "repeatedcv",
    number = 10,
    repeats = 5,
    verboseIter = TRUE
  )
)

# AD vs NPH
model_AD_NPH <- train(
  Pathology_group ~ .,
  AD_NPH,
  method = "ranger",
  tuneGrid = myGrid_AD_NPH,
  importance = "impurity",
  trControl = trainControl(
    method = "repeatedcv",
    number = 10,
    repeats = 5,
    verboseIter = TRUE
  )
)
```

### Top features of models
```{r}
importance_CU_AD <- as.data.frame(model_CU_AD[["finalModel"]]$variable.importance)
importance_CU_NPH <- as.data.frame(model_CU_NPH[["finalModel"]]$variable.importance)
importance_AD_NPH <- as.data.frame(model_AD_NPH[["finalModel"]]$variable.importance)

variable_importance <- data.frame(
  Features = rownames(importance_CU_NPH),
  CU_AD = importance_CU_AD$`model_CU_AD[["finalModel"]]$variable.importance`,
  CU_NPH = importance_CU_NPH$`model_CU_NPH[["finalModel"]]$variable.importance`,
  AD_NPH = importance_AD_NPH$`model_AD_NPH[["finalModel"]]$variable.importance`) %>%
  separate(Features, into = c("Gene_Name", "UniprotID"), sep = "_", remove = FALSE) %>%
  separate(Gene_Name, into = c("Razor_Protein", NA), sep = "\\.") %>%
  group_by(Razor_Protein) %>% 
  summarise(CU_AD = max(CU_AD), CU_NPH = max(CU_NPH), AD_NPH = max(AD_NPH)) %>%
  filter(!(Razor_Protein) %in% as.character(c(1:80)))
```

### Top features per contrast
```{r}
top_features_CU_AD <- variable_importance %>%
  select(Razor_Protein, CU_AD) %>%
  arrange(desc(CU_AD)) %>%
  filter(Razor_Protein != "Age" & Razor_Protein != "Gender") %>%
  slice(1:30)
top_features_CU_AD$Razor_Protein <- factor(top_features_CU_AD$Razor_Protein, levels = top_features_CU_AD$Razor_Protein[order(-top_features_CU_AD$CU_AD)])

CU_AD_features <- ggplot(top_features_CU_AD, aes(x= CU_AD, y = reorder(Razor_Protein, CU_AD))) +
  geom_bar(stat = "identity", position = "dodge", fill = "#646469") +
  labs(title = "CU vs AD", y = "Protein", x = "Gini importance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme_bw()
CU_AD_features

top_features_AD_NPH <- variable_importance %>%
  select(Razor_Protein, AD_NPH) %>%
  arrange(desc(AD_NPH)) %>%
  filter(Razor_Protein != "Age" & Razor_Protein != "Gender") %>%
  slice(1:30)
top_features_AD_NPH$Razor_Protein <- factor(top_features_AD_NPH$Razor_Protein, levels = top_features_AD_NPH$Razor_Protein[order(-top_features_AD_NPH$AD_NPH)])

AD_NPH_features <- ggplot(top_features_AD_NPH, aes(x= AD_NPH, y = reorder(Razor_Protein, AD_NPH))) +
  geom_bar(stat = "identity", position = "dodge", fill = "#646469") +
  labs(title = "AD vs NPH", y = "Protein", x = "Gini importance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme_bw()
AD_NPH_features

top_features_CU_NPH <- variable_importance %>%
  select(Razor_Protein, CU_NPH) %>%
  arrange(desc(CU_NPH)) %>%
  filter(Razor_Protein != "Age" & Razor_Protein != "Gender") %>%
  slice(1:30)
top_features_CU_NPH$Razor_Protein <- factor(top_features_CU_NPH$Razor_Protein, levels = top_features_CU_NPH$Razor_Protein[order(-top_features_CU_NPH$CU_NPH)])

CU_NPH_features <- ggplot(top_features_CU_NPH, aes(x= CU_NPH, y = reorder(Razor_Protein, CU_NPH))) +
  geom_bar(stat = "identity", position = "dodge", fill = "#646469") +
  labs(title = "CU vs NPH", y = "Protein", x = "Gini importance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme_bw()
CU_NPH_features

features_plot <- plot_grid(CU_AD_features, AD_NPH_features, CU_NPH_features, ncol = 3, align = "v")
features_plot
```

### Upset plot of top features
```{r}
Top_features_protein_names <- data.frame(CU_AD = top_features_CU_AD$Razor_Protein,
                                         AD_NPH = top_features_AD_NPH$Razor_Protein,
                                         CU_NPH = top_features_CU_NPH$Razor_Protein)
```

```{r}
upset_top_features_df <- data.frame(Proteins = unique(as.vector(as.matrix(Top_features_protein_names)))) %>%
  mutate(CU_AD = ifelse(Proteins %in% Top_features_protein_names$CU_AD, 1, 0)) %>%
  mutate(CU_NPH = ifelse(Proteins %in% Top_features_protein_names$CU_NPH, 1, 0)) %>%
  mutate(AD_NPH = ifelse(Proteins %in% Top_features_protein_names$AD_NPH, 1, 0))

upset(upset_top_features_df,
      order.by = "freq",
      text.scale = 1.4)

NPH_spec_top_proteins <- upset_top_features_df %>%
  filter(CU_NPH == 1 & AD_NPH == 1) %>%
  select(Proteins)
CU_spec_top_proteins <- upset_top_features_df %>%
  filter(CU_AD == 1 & CU_NPH == 1) %>%
  select(Proteins)
AD_spec_top_proteins <- upset_top_features_df %>%
  filter(CU_AD == 1 & AD_NPH == 1) %>%
  select(Proteins)

overlappers <- upset_top_features_df %>%
  mutate(sums = CU_AD + AD_NPH + CU_NPH) %>%
  filter(sums > 1)
```


### Heatmap of upset proteins
```{r}
heatmap_top_features <- protein_data_DEM_Apos %>%
  select(Participant, Pathology_group, c(6:ncol(protein_data_DEM_Apos))) %>%
  pivot_longer(3:ncol(.), names_to = "Protein_ID", values_to = "Abundance") %>%
  separate(Protein_ID, into = c("Protein_name", "Uniprot_ID"), sep = "_", remove = FALSE) %>%
  filter(Protein_name %in% upset_top_features_df$Proteins) %>%
  {df_top_proteins <<- .} %>%
  select(-c(Protein_ID, Uniprot_ID)) %>%
  unique() %>%
  group_by(Participant, Pathology_group, Protein_name) %>%
  dplyr::summarise(Abundance = mean(Abundance)) %>%
  pivot_wider(names_from = Protein_name, values_from = Abundance) %>%
  column_to_rownames("Participant")

heatmap_top_features_arranged <- arrange(heatmap_top_features, Pathology_group)

color <- list(Pathology_group = c("CU" = "#fee090", "AD" = "#4575b4", "NPH" = "#d73027"))
ha <- HeatmapAnnotation(Pathology_group = heatmap_top_features_arranged$Pathology_group, col = color)
heatmap <- Heatmap(t(heatmap_top_features_arranged[,2:ncol(heatmap_top_features_arranged)]),
        cluster_columns = TRUE,  # Enables clustering for columns
        cluster_rows = TRUE,    # Enables clustering for rows
        show_row_names = TRUE,  # Hides row names
        show_column_names = FALSE,  # Shows column names
        col = bluered(100),     # Sets the color palette
        top_annotation = ha,
        row_names_gp = gpar(fontsize = 10))
heatmap
```

### Boxplots of all top proteins
```{r}
top_features_boxplots <- heatmap_top_features_arranged %>%
  rownames_to_column("Participant") %>%
  pivot_longer(-c(Participant, Pathology_group), names_to = "Protein_ID", values_to = "Abundance") %>%
  separate(Protein_ID, into = c("Gene_symbol", "UniprotID"), sep = "_") %>%
  group_by(Participant, Pathology_group, Gene_symbol) %>%
  dplyr::summarise(Abundance = mean(Abundance)) %>%
  mutate(Pathology_group = factor(Pathology_group, levels = c("CU", "AD", "NPH"))) %>%
  mutate(Gene_symbol = factor(Gene_symbol, levels = c("BASP1", "MDH1", "GOT1", "PKM", "PGLYRP2", "F5", "LTBP2", "KIAA1549L", "VGF", "NPTXR", "NRXN3", "PTPRN2"))) %>%
  filter(!is.na(Gene_symbol))

color <- c("AD" = "#4575b4", "NPH" = "#d73027", "CU" = "#fee090")
ggplot(top_features_boxplots, aes(x = Pathology_group, y = Abundance, fill = Pathology_group)) +
  geom_boxplot() +
  scale_fill_manual(values = color) +
  facet_wrap(~Gene_symbol, ncol = 6)+
  labs(x = "Group") +
  theme_bw() +
  theme(strip.text = element_text(size=10))
```

# ROC curves top proteins
### AD vs NPH
```{r}
toROC_AD_NPH <- protein_data_DEM_Apos %>%
  select(Pathology_group,BASP1_P80723,MDH1_P40925, GOT1_P17174, PKM_P14618, PGLYRP2_Q96PD5, F5_P12259) %>%
  filter(Pathology_group != "CU") %>%
  mutate(Pathology_group = ifelse(Pathology_group == "NPH", 1, 0))

# One simple ROC
roc_BASP1 <- roc(toROC_AD_NPH$Pathology_group, toROC_AD_NPH$BASP1_P80723)
roc_MDH1 <- roc(toROC_AD_NPH$Pathology_group, toROC_AD_NPH$MDH1_P40925)
roc_GOT1 <- roc(toROC_AD_NPH$Pathology_group, toROC_AD_NPH$GOT1_P17174)
roc_PKM <- roc(toROC_AD_NPH$Pathology_group, toROC_AD_NPH$PKM_P14618)
roc_PGLYRP2 <- roc(toROC_AD_NPH$Pathology_group, toROC_AD_NPH$PGLYRP2_Q96PD5)
roc_F5 <- roc(toROC_AD_NPH$Pathology_group, toROC_AD_NPH$F5_P12259)

# Using the glm to create composite score across all top proteins
model <- glm(Pathology_group ~ ., data = toROC_AD_NPH, family = "binomial")
toROC_AD_NPH$glm_res <- predict(model, type = "response")

glm_roc_object_AD_NPH <- roc(toROC_AD_NPH$Pathology_group, toROC_AD_NPH$glm_res)
auc(glm_roc_object_AD_NPH)

# combine results
roc_results <- list(roc_BASP1, roc_MDH1, roc_GOT1, roc_PKM, roc_PGLYRP2, roc_F5, glm_roc_object_AD_NPH)
names(roc_results) <- c("BASP1", "MDH1", "GOT1", "PKM", "PGLYRP2", "F5", "Composite")

# plot
ggroc(roc_results) +
  geom_segment(aes(x = 0, xend = 1, y = 1, yend = 0), linetype = "dashed", color = "darkgrey") +
  scale_color_manual(values = c("BASP1" = "red", "MDH1" = "green", "GOT1" = "blue", "PKM" = "orange", 
                                 "PGLYRP2" = "purple", "F5" = "brown", "Composite" = "black")) +
  theme_bw() +
  ggtitle("AD vs NPH")
```

### CU vs NPH
```{r}
toROC_CU_NPH <- protein_data_DEM_Apos %>%
  select(Pathology_group,LTBP2_Q14767,KIAA1549L_H0YDE5, VGF_O15240, NPTXR_O95502, NRXN3_Q9Y4C0, PTPRN2_Q92932) %>%
  filter(Pathology_group != "AD") %>%
  mutate(Pathology_group = ifelse(Pathology_group == "NPH", 1, 0))

# One simple ROC
roc_LTBP2 <- roc(toROC_CU_NPH$Pathology_group, toROC_CU_NPH$LTBP2_Q14767)
roc_KIAA1549 <- roc(toROC_CU_NPH$Pathology_group, toROC_CU_NPH$KIAA1549L_H0YDE5)
roc_VGF <- roc(toROC_CU_NPH$Pathology_group, toROC_CU_NPH$VGF_O15240)
roc_NPTXR <- roc(toROC_CU_NPH$Pathology_group, toROC_CU_NPH$NPTXR_O95502)
roc_NRXN3 <- roc(toROC_CU_NPH$Pathology_group, toROC_CU_NPH$NRXN3_Q9Y4C0)
roc_PTPRN2 <- roc(toROC_CU_NPH$Pathology_group, toROC_CU_NPH$PTPRN2_Q92932)

# Using the glm to create composite score across all top proteins
model <- glm(Pathology_group ~ ., data = toROC_CU_NPH, family = "binomial")
toROC_CU_NPH$glm_res <- predict(model, type = "response")

glm_roc_object_CU_NPH <- roc(toROC_CU_NPH$Pathology_group, toROC_CU_NPH$glm_res)
auc(glm_roc_object_CU_NPH)

# combine results
roc_results <- list(roc_LTBP2, roc_KIAA1549, roc_VGF, roc_NPTXR, roc_NRXN3, roc_PTPRN2, glm_roc_object_CU_NPH)
names(roc_results) <- c("LTBP2", "KIAA1549", "VGF", "NPTXR", "NRXN3", "PTPRN2", "Composite")

# plot
ggroc(roc_results) +
  geom_segment(aes(x = 0, xend = 1, y = 1, yend = 0), linetype = "dashed", color = "darkgrey") +
  scale_color_manual(values = c("LTBP2" = "red", "KIAA1549" = "green", "VGF" = "blue", "NPTXR" = "orange", 
                                 "NRXN3" = "purple", "PTPRN2" = "brown", "Composite" = "black")) +
  theme_bw() +
  ggtitle("CU vs NPH")
```

### CU vs AD
```{r}
toROC_CU_AD <- protein_data_DEM_Apos %>%
  select(Pathology_group,SMOC1_Q9H4F8,PGK1_P00558, LDHA_P00338, ENO1_P06733, CALM1.CALM2.CALM3_P0DP23.P0DP24.P0DP25, ALDOA_P04075) %>%
  filter(Pathology_group != "NPH") %>%
  mutate(Pathology_group = ifelse(Pathology_group == "AD", 1, 0))

# One simple ROC
roc_SMOC1 <- roc(toROC_CU_AD$Pathology_group, toROC_CU_AD$SMOC1_Q9H4F8)
roc_PGK1 <- roc(toROC_CU_AD$Pathology_group, toROC_CU_AD$PGK1_P00558)
roc_LDHA <- roc(toROC_CU_AD$Pathology_group, toROC_CU_AD$LDHA_P00338)
roc_ENO1 <- roc(toROC_CU_AD$Pathology_group, toROC_CU_AD$ENO1_P06733)
roc_CALM1 <- roc(toROC_CU_AD$Pathology_group, toROC_CU_AD$CALM1.CALM2.CALM3_P0DP23.P0DP24.P0DP25)
roc_ALDOA <- roc(toROC_CU_AD$Pathology_group, toROC_CU_AD$ALDOA_P04075)

# Using the glm to create composite score across all top proteins
model <- glm(Pathology_group ~ ., data = toROC_CU_AD, family = "binomial")
toROC_CU_AD$glm_res <- predict(model, type = "response")

glm_roc_object_CU_AD <- roc(toROC_CU_AD$Pathology_group, toROC_CU_AD$glm_res)
auc(glm_roc_object_CU_AD)

# combine results
roc_results <- list(roc_SMOC1, roc_PGK1, roc_LDHA, roc_ENO1, roc_CALM1, roc_ALDOA, glm_roc_object_CU_AD)
names(roc_results) <- c("SMOC1", "PGK1", "LDHA", "ENO1", "CALM1", "ALDOA", "Composite")

# plot
ggroc(roc_results) +
  geom_segment(aes(x = 0, xend = 1, y = 1, yend = 0), linetype = "dashed", color = "darkgrey") +
  scale_color_manual(values = c("SMOC1" = "red", "PGK1" = "green", "LDHA" = "blue", "ENO1" = "orange", 
                                 "CALM1" = "purple", "ALDOA" = "brown", "Composite" = "black")) +
  theme_bw() +
  ggtitle("CU vs AD")
```

