---
title: "GSEA"
author: "Matthijs de Geus"
output:
  html_document:
    toc: yes
    toc_float: yes
    code_folding: show
  '': default
---

### Load data and dependencies
```{r}
library(tidyverse)
library(janitor)
library(gplots)
protein_data <- read.csv("protein_data.csv")
string_db_res <- read.csv("lm_GSEA_results_combined.csv")
```

### restructuring protein data df
Columns for each pathology diagnosis with the average abundance for each protein. Then add columns for each contrast.
```{r}
fold_change_data <- protein_data_DEM_Apos %>%
  select(Pathology_group, c(55:ncol(protein_data_DEM_Apos))) %>%
  group_by(Pathology_group) %>%
  summarise(across(everything(), mean)) %>%
  t() %>%
  as.data.frame() %>%
  row_to_names(row_number = 1) %>%
  rownames_to_column("Protein") %>%
  {. ->> avg_protein_levels_pathology} %>%
  mutate(AD_vs_CU = (as.numeric(AD)-as.numeric(CU)),
         AD_vs_NPH = (as.numeric(AD)-as.numeric(NPH)),
         CU_vs_NPH = (as.numeric(CU)-as.numeric(NPH))) %>%
  select(Protein, AD_vs_CU, AD_vs_NPH, CU_vs_NPH) %>%
  separate(Protein, into = c("Protein_name", "UniprotID"), sep = "_") %>%
  separate(UniprotID, into = c("UniprotID", NA), sep = "\\.")
```


# Upset plot of gene sets
#### Create the Upset data frame
```{r}
CU_AD_pathways <- string_db_res %>%
  filter(Contrast == "CU_AD") 
  
AD_NPH_pathways <- string_db_res %>%
  filter(Contrast == "AD_NPH")
  
CU_NPH_pathways <- string_db_res %>%
  filter(Contrast == "CU_NPH") 

upset_df <- data.frame(Pathways = c(as.vector(unique(string_db_res$term.description)))) %>% 
  mutate(`CU vs AD` = ifelse(Pathways %in% CU_AD_pathways$term.description, 1, 0)) %>%
  mutate(`AD vs NPH` = ifelse(Pathways %in% AD_NPH_pathways$term.description, 1, 0)) %>%
  mutate(`CU vs NPH` = ifelse(Pathways %in% CU_NPH_pathways$term.description, 1, 0)) %>%
  column_to_rownames(var = "Pathways") %>%
  mutate(sums = rowSums(.[1:3]))
```

#### Upset plot
```{r}
UpSetR::upset(upset_df, 
              order.by = c("degree"),
              # group.by = "degree",
              decreasing = FALSE,
              # intersections = intersections_list,
              matrix.color = "grey20",
              main.bar.color = "grey10",
              shade.color = "grey70",
              sets.x.label = "",
              # number.angles = 45,
              mainbar.y.label = "Number of Pathways",
              text.scale = 1.4
              )
```


## Extracting the proteins for heatmap
```{r}
selected_pathways <- string_db_res %>%
  filter(term.description %in% rownames(upset_df)) %>%
  select(term.description, matching.proteins.in.your.input..labels.) %>%
  separate_rows(matching.proteins.in.your.input..labels., sep = ",") %>%
  rename("Protein_name" = matching.proteins.in.your.input..labels.)

protein_data_long <- protein_data_DEM_Apos %>%
  pivot_longer(-c(1:55), names_to = "Protein_ID", values_to = "Abundance") %>%
  separate(Protein_ID, into = c("Protein_name", "UniprotID"), sep = "_", remove = FALSE) %>%
  separate(Protein_name, into = c("Protein_name", NA), sep = "\\.")

grouped_protein_data_long <- protein_data_long %>%
  group_by(Participant, Pathology_group, Age, Protein_name) %>%
  summarise(Goup_abundance_max = mean(Abundance)) %>%
  filter(Protein_name %in% selected_pathways$Protein_name) %>%
  filter(Pathology_group != "Other")

merged_df <- merge(selected_pathways, grouped_protein_data_long) %>%
  group_by(Participant, Pathology_group, term.description) %>%
  summarise(pathway_abundance = mean(Goup_abundance_max))
```

#### proteins to highlight in heatmap for specific pathways
```{r}
glucose_module <- c("Glycolytic process",
                    "Generation of precursor metabolites and energy",
                    "Nucleobase-containing small molecule metabolic process",
                    "Nucleotide metabolic process",
                    "Purine ribonucleotide metabolic process",
                    "Small molecule catabolic process")
glucose_module_proteins <- selected_pathways %>%
  filter(term.description %in% glucose_module) 

axon_module <- c("Axon guidance",
                 "Axonogenesis",
                 "Cell morphogenesis involved in neuron differentiation",
                 "Cell projection morphogenesis",
                 "Neuron projection morphogenesis",
                 "Cellular component morphogenesis")

axon_module_2 <- c("Axon development",
                 "Plasma membrane bounded cell projection organization",
                 "Neuron projectino development",
                 "Neuron development",
                 "Generation of neurons",
                 "Neuron differentiation")

axon_module_proteins <- selected_pathways %>%
  filter(term.description %in% axon_module_2 | term.description %in% axon_module) 
axon_module_prot_count <- as.data.frame(table(axon_module_proteins$Protein_name)) %>%
  filter(Freq > 30)

complement_module <- c("Complement activation, classical pathway",
                       "Immunoglobulin mediated immune response",
                       "Complement activation")
complement_module_proteins <- selected_pathways %>% filter(term.description %in% complement_module)
complement_module_prot_count <- as.data.frame(table(complement_module_proteins$Protein_name)) %>%
  filter(Freq > 3)


vascular_module <- c("Negative regulation of wound healing",
                     "Negative regulation of coagulation",
                     "Negative regulation of blood coagulation",
                     "Regulation of blood coagulation")
vascular_module_proteins <- selected_pathways %>% filter(term.description %in% vascular_module)
vascular_module_prot_count <- as.data.frame(table(vascular_module_proteins$Protein_name)) %>%
  filter(Freq > 3)
```


## Now plot the heatmap of the abundances
```{r}
## Make the heatmap matrices
## CU AD heatmap
CU_AD_heatmap <- merged_df %>%
  filter(Pathology_group == "AD" | Pathology_group == "CU") %>%
  filter(term.description %in% CU_AD_pathways$term.description) %>%
  pivot_wider(names_from = term.description, values_from = pathway_abundance) %>%
  column_to_rownames("Participant")

## AD NPH heatmap
AD_NPH_heatmap <- merged_df %>%
  filter(Pathology_group == "AD" | Pathology_group == "NPH") %>%
  filter(term.description %in% AD_NPH_pathways$term.description) %>%
  pivot_wider(names_from = term.description, values_from = pathway_abundance) %>%
  column_to_rownames("Participant") 

## CU NPH heatmap
CU_NPH_heatmap <- merged_df %>%
  filter(Pathology_group == "NPH" | Pathology_group == "CU") %>%
  filter(term.description %in% CU_NPH_pathways$term.description) %>%
  pivot_wider(names_from = term.description, values_from = pathway_abundance) %>%
  column_to_rownames("Participant")

## Make the heatmaps
color <- list(Pathology_group = c("CU" = "#fee090", "AD" = "#4575b4"))
ha <- HeatmapAnnotation(Pathology_group = CU_AD_heatmap$Pathology_group, col = color)
heatmap_CU_AD <- Heatmap(t(CU_AD_heatmap[,2:ncol(CU_AD_heatmap)]),
        cluster_columns = TRUE,  # Enables clustering for columns
        cluster_rows = TRUE,    # Enables clustering for rows
        show_row_names = TRUE,  # Hides row names
        show_column_names = FALSE,  # Shows column names
        col = bluered(100),     # Sets the color palette
        top_annotation = ha,
        row_names_gp = gpar(fontsize = 7)
)
draw(heatmap_CU_AD)

color <- list(Pathology_group = c("CU" = "#fee090", "NPH" = "#d73027"))
ha <- HeatmapAnnotation(Pathology_group = CU_NPH_heatmap$Pathology_group, col = color)
heatmap_CU_NPH <- Heatmap(t(CU_NPH_heatmap[,2:ncol(CU_NPH_heatmap)]),
        cluster_columns = TRUE,  # Enables clustering for columns
        cluster_rows = TRUE,    # Enables clustering for rows
        show_row_names = TRUE,  # Hides row names
        show_column_names = FALSE,  # Shows column names
        col = bluered(100),     # Sets the color palette
        top_annotation = ha,
        row_names_gp = gpar(fontsize = 7)
)
draw(heatmap_CU_NPH)


color <- list(Pathology_group = c("AD" = "#4575b4", "NPH" = "#d73027"))
ha <- HeatmapAnnotation(Pathology_group = AD_NPH_heatmap$Pathology_group, col = color)
heatmap_AD_NPH <- Heatmap(t(AD_NPH_heatmap[,2:ncol(AD_NPH_heatmap)]),
        cluster_columns = TRUE,  # Enables clustering for columns
        cluster_rows = TRUE,    # Enables clustering for rows
        show_row_names = TRUE,  # Hides row names
        show_column_names = FALSE,  # Shows column names
        col = bluered(100),     # Sets the color palette
        top_annotation = ha,
        row_names_gp = gpar(fontsize = 7)
)
draw(heatmap_AD_NPH)
```

Heatmap of overlapping pathways only
```{r}
three_pathways <- upset_df %>%
  filter(sums > 1)

Overlapping_pathways <- merged_df %>%
  filter(term.description %in% rownames(three_pathways))

Overlapping_pathways_heatmap <- Overlapping_pathways %>%
  pivot_wider(names_from = term.description, values_from = pathway_abundance) %>%
  column_to_rownames("Participant")

color <- list(Pathology_group = c("CU" = "#fee090", "NPH" = "#d73027", "AD" = "#4575b4"))
ha <- HeatmapAnnotation(Pathology_group = Overlapping_pathways_heatmap$Pathology_group, col = color)

Pathways_overlap_heatmap <- Heatmap(t(Overlapping_pathways_heatmap[,2:ncol(Overlapping_pathways_heatmap)]),
        cluster_columns = TRUE,  # Enables clustering for columns
        cluster_rows = TRUE,    # Enables clustering for rows
        show_row_names = TRUE,  # Hides row names
        show_column_names = FALSE,  # Shows column names
        col = bluered(100),     # Sets the color palette
        top_annotation = ha,
        row_names_gp = gpar(fontsize = 6)
)
draw(Pathways_overlap_heatmap)
```
